% !TeX root = ../../thesis.tex
\chapter*{Personal Acknowledgments}                                  \label{ch:preface}

In a 1991 interview with Academy of Achievement, evolutionary biologist Stephen Jay Gould said ``It'd be nice if we could claim that anyone who's driven enough will rise, but that's not true. Luck has to play an enormous role in anyone's career.''
As I reach the point in my own career where I am able to write the personal acknowledgments to my doctoral thesis I am overwhelmed by the feeling of how lucky I have been to get to this stage.
Of course, luck comes from may sources: as Dr. Gould pointed out, I am lucky to have been born in a time and place in history where I was not compromised by any early-childhood diseases, and I had freedom to pursue an academic career path without restriction.
Throughout my whole life I have been fortunate to be able to attend some of the best quality institutions and receive some of the best quality education available.
Above any material conditions, however, I feel most lucky to have met and been lifted up by an extraordinary number of people who have made getting to this culmination of almost three decades of formal education possible.
I know that I cannot possibly thank everyone, and that a few words on a page are insufficient for the contributions made by those listed here.

Still, I do need to say a sincere ``thank you'' to some people.

First, I want to thank my entire doctoral jury, beginning with my external jury members, Drs. Verity Hill and Bram Vrancken, for your thoughtful and thorough reviews of this thesis, as well as your willingness to examine me at my doctoral defense---my thesis and knowledge are both much stronger for your contributions.
For the same reasons, but also for your repeated feedback, support, and advice since the very beginning of my doctoral work I want to thank Profs. Jelle Matthijnssens and Philippe Lemey.
I cannot understate how much your guidance has helped steer me towards success during my PhD.
Finally, thank you Prof. Kristel Van Laethem, for serving as the chair of my doctoral defense.

My doctoral thesis would not have been possible without the support and direction of my promotor.
Thank you, Guy, for patiently teaching me so much about phylodynamics, Bayesian inference, and generally showing me how to perform high-quality research.
Thank you for taking a risk on a predoctoral student who had never set foot in Belgium and for believing in me and supporting me through every step of this process---it hasn't always been easy, but I'm absolutely sure I could not have made it through this process without your guidance and expertise.
Also, thanks for all the cinnamon rolls.
They gave me a taste of home when I needed it most.

To my co-promotor, Mandev, I also need to give my heartfelt thanks.
From the beginning of my predoc you've been a supportive and joyful presence.
Thank you for helping me adapt to life in Belgium at the beginning of this whole journey, and thank you even more for all of your academic guidance, feedback on manuscripts, and patient teaching when I simply could not grasp the Hastings ratio.
I look forward to being able to collaborate a bit from the same time zone in the future.

Of course, my time working at the Rega Institute was brightened by a whole cast of office- and lab-mates who not only taught me so much about science, but also made the proposition of going to work amidst the most disruptive event in a hundred years while living in a brand new country exciting.
I would like to give specific thanks to three office-mates who have been there since the start: Kanika, thank you for bringing laughs, thoughtful conversation, and perspective to the past five years, both in the office and out; Sam, thank you for always being the PhD big brother that I could look up to for just about every question and problem I ever ran into; Ine, not a single day has gone by in the office that you didn't go out of your way to make my life (and everyone else's) easier, you da real MVP.
Thank you to the other Baele Lab members who have brought both thoughtful scientific feedback and so much joy (and a bit of weird): Marius, Nena, and Yimin.
Thanks to all of the ECV lab members past and present who have not only been excellent colleagues, but also amazing friends: Emanuele, Fre, Jade, Joon, Liana, Magda, Senne, Sophie, Yiqiao, and too many visiting scholars and masters students to name.
Finally, I want to acknowledge Cathy De Meyer for working tirelessly to make any issue that came up over the past few years disappear with ease.

Even before I arrived in Belgium for this PhD I received mentorship and support from academics and teachers without whom I am sure I wouldn't have made it to this point.
Thank you to Trevor Bedford for getting me started in viral phylogenetics in the first place, I owe so much of the successes that I've had to your mentorship.
Additionally, a huge thank you to the blab members who taught me so much of the foundational skills that I learned during that time.
In particular, I'd like to thank Alli, James, John, Sidney, and Tom.
Thank you to Anna Ritz and Jim Fix for getting me started down the path of computational biology, and thank you to Sarah Schaack and Jeremy Coate for teaching me that research science is a career path that I love---the four of you are the real answer to the question, ``Why Reed?''
Finally, thank you to Roy Ledesma for teaching me to love mathematics, and to Shirley Rau for teaching me how to write.

Not only have I been incredibly lucky to have excellent colleagues, but I have also been extremely fortunate to have a vast support network of friends who have kept me alive, sane, and occasionally happy throughout my doctoral studies.
I'd like to take a moment to thank some of you (alphabetically).
Andrea, my Captain, thank you for being an okay Magic player, a good-music enjoyer, a great pitch blocker, and an incredible friend; you've always managed to be even more than I've needed you to be.
Caleb, thank you for being a rock in my life ever since that first Streetlight show; you're the best DM I've ever had.
Colten, my first best friend, thank you for 25 years of being the person I can always talk to about anything as if no time has passed.
David, thank you for teaching me how to enjoy life in Belgium and for making me feel welcomed when no one else did.
Emma, thank you for countless video calls during the lockdowns and for so so many animal stories that should not be repeated.
Gytis, I wasn't sure which section of these acknowledgments I should list you in because of how many parts of my life you have touched; thank you so much for being both a mentor and a great friend since the day we first met in Seattle.
Karan, thank you for always being a shining example of what it means to be a kind and supportive friend.
Melanie, thank you for guiding me towards my next big adventure; I can't wait to be colleagues.
Michael, I could thank you for so many things here, but in this context thank you for teaching me what it means to be a passionate and driven person who puts love into everything they do, both work and personal.
Oliver, thank you for being the spark that re-ignited my love of baseball.
Rema, thank you for teaching me that putting the hard work into friendships is always worth the effort.
Toine, thank you for every minute of love and support you've poured into the last few years, and for demonstrating how to be my authentic self; also, thank you for only catching once and otherwise sticking to being a rock at first.
Finally, Vanessza, thank you so much for patiently and caringly supporting me through the rockiest period of my doctoral studies; I don't think I would have made it over the finish line without you.

When I first moved to Belgium I didn't have any friends or hobbies in this continent.
Fortunately for me, both of these issues were quickly solved when I showed up for try-outs for my baseball team, the Leuven Twins.
I have been touched by so many ballplayers through my involvement with the Twins, so I can't hope to list everyone.
However, I will thank a few of my teammates and friends from the team to whom I owe years of fun and out-of-work purpose.
In particular, I'd like to thank Arno, Daan, Dima, Freek, John, Masa, Omar, PJ, Will, and Yosuke for always picking me up when I needed it, both on the field and off---thank you all supporting me through all my temperamental moments and blow-up starts.
I also want to thank Aila, Lea, and Silke from the Twins for being my board game buds.
During my doctoral studies all of you on the Twins have been my family-within-Belgium.

Finally, I have to thank the group of people who have undoubtedly given me the most love and support through this PhD, and through my whole life.
Mom, I know I can't begin to thank you in this paragraph for everything that you've done for me.
Thank you for always giving me a parent in school to look up to, for making sure I got to North Star, for always giving me sound advice and unconditional love, for all the nicknames, and for teaching me how to use language like a paintbrush.
I'm sorry that I couldn't fit a seal into this thesis.
Dad, thank you for always pushing me to be better than what I thought I could be, for teaching me to love nature and science from a young age, for always being present and being an advocate for me, Sam, and all our classmates when no one else in Emmett would.
Also, thanks for painting such great cover art!
Sam, I couldn't ask for more in a brother.
You are my role model for how to be a thoughtful, curious, and principled person.
I'm so proud to be your brother.
Naomi, thank you for always being a down-to-earth and rational force in the family when we so desperately need it.
Lastly, I'd like to thank all four of my grandparents, even though none of you are able to see this day you all have touched me so deeply that there isn't a day I don't think about you: Grandma Potter, thank you for teaching me love; Grandpa Potter, thank you for your memories and your memory; Grandma Carol, thank you for teaching me to find joy in another person's existence; Far Far, thank you for teaching me to laugh and for being my best friend for so many years.
I love all of you; thank you for everything, family.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Keep the following \cleardoublepage at the end of this file, 
% otherwise \includeonly includes empty pages.
\cleardoublepage

% vim: tw=70 nocindent expandtab foldmethod=marker foldmarker={{{}{,}{}}}
